---
title: "Class 5: Data Viz with ggplot"
author: "Laura (A17844089)"
format: pdf 
---

Today we are exploring the **gggplot** package and how to make nice figures in R. 

There are lots of ways to make figures and plot in R. Theses include: 

- so called "base" R 
- and add on package like **ggplot2** 

Here is a simple "base" R plot.


```{r}
head(cars)
```

We can simply pass to to the `plot()` function. 
```{r}
plot(cars)
```
```{r}

```

Key-point: Base R is quick but not so nice and simple looking in some folks eyes 

Let's see how we can plot this with **ggplot2**... 

1st I need to install this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal. 

2nd We need to load the package with the `library()` function every time we want to use it. 

```{r}
library(ggplot2)
ggplot(cars)
```


```{r}
ggplot(cars)
``` 


Every ggplot is composed of at least 3 layers: 

- **data** (i.e a data.frame with the things you want to plot), 
- aesthitics **aes()** that map the columns of data to your plot features (i.e aesthitics) 
- geoms like **geom_point)** that srt how the plot appears 

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point()
```

```{r}
hist(cars$speed)
```

> Key point: For simple "canned" graphs base R is quicker but as things get more custom and elobrate then ggplot wins out...


Let's add more layers to our ggplot 

Add a line showing the relationship between x and y 
Add a title 
Add custom axis labels "Speed (MPH)" and "Distance (ft)" 
Change the theme...

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) + 
 labs(title = "Silly plot of Speed vs Stoping distance", 
      x="Speed (MPH)", 
      y="Distance (ft)") + 
  theme_bw()
```

## Going further 

Read some gene expression data 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset? 

```{r}
nrow(genes) 
```


> Q2. How many "up" regulated genes are there? 

```{r}
sum( genes$State =="up" ) 
``` 

A useful function for counting up occurances of things in a vector is the `table()` function. 
Make a v1 figure 

```{r}
p <- ggplot(genes) + 
  aes(x=Condition1,
      y=Condition2,  
       col=State) + 
  geom_point()
p
```
```{r}
p + 
 scale_colour_manual( values= c("blue","gray","red") ) + 
  labs(title= "Expression changes upon drug treatment", 
       x="Control (no drug)",
       y= "Treatment (with drug)" ) + 
    theme_bw() 
```



```{r}
# File location online 
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url) 
```

Lets have a wee peak 

```{r} 
head( gapminder, 3) 
```


> Q4. How many different country values are in this dataset? 

```{r}
head( gapminder, 3) 
```

```{r}       
length( table(gapminder$country) ) 
```

>Q5. How many different continent values are in the dataset?  

```{r}
unique(gapminder$continenet) 
```

```{r}
 ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent) + 
  geom_point()
```

```{r}
 ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent, label=country) + geom_point() 
  geom_point()
```       
       
I can use the **ggrepl** package to make more sensible labels here. 

I want a seperate panel per continent 

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent, label=country) + geom_point() + 
facet_wrap(~continent) 
```   

```{r}
library(ggrepel)
       
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent, label=country) + geom_point() 
```    

The advantages of ggplot over base R plot: 
Layered system: ggplot builds plots by adding layers (data, aesthetics, geometry, themes, annotations) with the + operator, making it easy to iteratively refine and customize visualizations. Base R requires step-by-step specification, which can be more fiddly and time-consuming to polish
D1
D2
D4
D5
D6
Q2
.
Aesthetic mapping: ggplot allows mapping data columns to visual features like position, color, shape, size, and transparency, supporting rich, multi-dimensional plots. Base R is less flexible in this regard
D1
D2
D3
D5
D6
Q1
Q3
.
Themes and annotation: ggplot provides built-in themes and functions for adding titles, subtitles, captions, and custom axis labels, making it easier to standardize and polish figures
D1
D2
D4
D5
D6
Q1
.
Faceting: ggplot can split data into multiple panels by categories using facet_wrap, improving clarity for complex or grouped data
D1
D2
D4
D5
D6
Q1
.
Variety of plot types: ggplot supports over 40 core geometric types and many more via extensions, making advanced visualizations more accessible
D1
D2
D6
Q3
.
In summary, ggplot is more structured, modular, and user-friendly for creating publication-quality figures, while base R offers complete control but is less convenient for complex or polished visualizations
D1
D2
D4
D5
D6
Q1
Q2
Q3
.









   