---
title: "Class 8: Breast Cancer Analysis Mini Project"
author: "Laura (PID: A17844089)"
format: pdf
toc: true 
---

## Background 
The goal of this mini-project is to explore a complete analysis using the unsupervised learning techniques covered in our class. 

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data import 

Data was downloaded from the class website as a CSV file. 

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1) 
head(wisc.df) 
``` 

## Data Exploration 

The first column `diagonosis` is the expert opinion on the sample (i.e. patient FNA) 


```{r}
head(wisc.df$diagnosis) 
```

```{r}
head(wisc.df[,-1]) 
```

Remove the diagnosis from data for subsequent analysis 
```{r}
wisc.data <- wisc.df[,-1] 
dim(wisc.data) 
```

Store the diagnosis as a vector for use later when we compare our results to those from experts in the field. 

```{r}
diagnosis <- factor(wisc.df$diagnosis)
```

> Q1. How many observations are in this dataset? 

There are `r nrow(wisc.data)` observations/patients in the dataset 

> Q2. How many of the observations have a malignant diagnosis? 

```{r}
table(wisc.df$diagnosis) 
```


> Q3. How many variables/features in the data are suffixed with _mean? 

```{r}
colnames(wisc.data)
```

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)) ) 
```


## Principal Component Analysis (PCA) 

The `prcomp()` function to do PCA has a `scale=FALSE` default. In general we nearly always want to set this to TRUE so our analysis is not dominated by columns/variables in our dataset that have high standard deviation and mean when compared to others just because the units of measurement are on different scales. 


```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE) 
summary(wisc.pr) 
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)? 

Proportion of Variance (PC1) = 0.4427 


> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data? 

PC1: 0.4427
PC2: 0.6324 
PC3: 0.72636 

Three principal componets (PC1-PC3) are required to explain at least 70% of the total variance. 


>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data? 

PC6: 0.88759
PC7: 0.91010

Seven principal components (PC1-PC7) are required to explain at least 90% of the total variance. 



### PCA Score Plot 
The main PC result figure is called a "score plot" or "PC plot" or "ordination plot"... 

```{r}
library(ggplot2)

ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point() 
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why? 

The plot looks very cluttered and close together because the dataset has many observations and features. The overlapping points make it difficult to distinguish individual samples or patterns. 


```{r}
library(ggplot2)

ggplot(wisc.pr$x) + 
  aes(PC1, PC3, col=diagnosis) + 
  geom_point() 
```
> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots? 

The separation between groups is less distinct. Most of the variation is still along PC1, while PC3 adds less new information and doesn't separate the data into clusters. 




## PCA Screeplot 

A plot of how much variance each PC captures. We can get this from `wisc.pr$sdev` or from the output of `summary(wisc.pr)` 


```{r}
wisc.pr$sdev 
```

```{r}
var.tbl <- summary(wisc.pr) 
head(var.tbl$importance)
```

```{r}
var <- var.tbl$importance[2,]
cum.var <- var.tbl$importance[3,] 

plot(var * 100, typ="b", 
     ylab="Percent Variance Captured", 
     xlab= "PC Number") 
```


```{r}

plot(cum.var * 100, typ="o", 
     ylab="Percent Variance Captured", 
     xlab= "PC Number") 
ylim=c(0,100)
points(var * 100, col="blue", type="o") 
```

### Communicating PCA results 

> Q9. For the first principal component, what is the component of the loading vecter (i.e wisc.pr$rotation[,1]) for the feature concave.points_mean? 

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```


> Q10. What is the minimum number of principal componenets required to explain 80% of the variance of the data? 


We need 5 PCs to capture more than 80% variance 

```{r}

summary(wisc.pr) 
```

## Hierarchical clustering 

Just clustering the original data is not very informative or helpful. 

```{r} 
data.scaled <- scale(wisc.data) 
data.dist <- dist(data.scaled) 
wisc.hclust <- hclust(data.dist) 

```

View the clustering dendrogram result 

```{r}
plot(wisc.hclust) 
```



```{r} 
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters) 

```

```{r}
table(wisc.hclust.clusters, diagnosis) 
```


## 5. Combining methods (PCA and Clustering) 

Clustering the original data was not very productive. The PCA results looked promising. Here we combine these methods by clustering from our PCA results. In other words "clustering in PC space"... 

```{r} 
## Take the first 3 PCs
dist.pc <- dist( wisc.pr$x[, 1:3] ) 
wisc.pr.hclust <- hclust(dist.pc, method="ward.D2") 

```


View the tree... 
```{r}
plot(wisc.pr.hclust)
abline(h=70, col="red") 
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters? 

The height in the plot is at 70. 







To get one clustering membership vector (i.e. our main clustering result) we can see the tree at a desired height or to yield a deserved number of "k" groups 

```{r}
grps <- cutree(wisc.pr.hclust, h=70) 
table(grps) 
``` 


How does this clustering groups compare to the expert 

```{r} 
table(grps, diagnosis)
```




> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10? 

The best alignment with the actual diagnoses was when the tree was cut to yield two clusters. Using more than two clusters tended to split groups without helping the separation between diagnoses. 

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

Among the methods tested, the Ward.D2 method gave the clearest and most distinct clusters. Ward.D2 effectively grouped samples into clusters that aligned closely with the diagnoses. 

> Q15. How well does the newly created model with four clusters separate out the two diagnoses? 

The hierarchical clustering model and the Ward.D2 method produces two clusters. The data improved the clustering results compared to the model using the original scaled data. 



```{r}
# Compare hierarchical clustering (before PCA) to actual diagnoses
table(wisc.hclust.clusters, diagnosis)
```

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses. 

The k-means model results show a moderate overlap between diagnoses while the hierarchical clustering model without PCA produced four clusters that only partially corresponded to the actual diagnoses. After applying PCA, the plot had more separation and helped remove noise features. I think hierarchical clustering on PCA reduced data provided the best match. 

## 6. Sensitivity/Specificity 

Sensitivity: TP/(TP+FN) 
Specificity: TN/(TN+FN) 

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Among all of the methods, the hierarchical clustering after PCA achieved the best balance between sensitivity and specificity. In contrast, both the hierarchical clustering before PCA and k means clustering models had lower sensitivity and specificity, most likely due to overlapping clusters and correlated features. 


## 7. Prediction 

We can use our PCA model for prediction with new input patient samples.  

> Q18. Which of these new patients should we prioritize for follow up based on your results? 

Based on the plot, Patient 1 should be prioritized for a follow up and diagnostic testing because their PCA pattern follows the previous diagnosed malignant cases. 





