---
title: "Class 7: Machine Learning 1"
author: "Laura (PID: 17844089)"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction. 

## K-means clustering 

To see how this works let's first makeup some data to cluster where we know what the answer should be. We can use the `rnorm()` function to help here: 

```{r}
hist( rnorm(500, mean=5) ) 
```


```{r}
x <- c( rnorm(30, mean=-3), rnorm(30, mean=3) ) 
y <- rev(x)
```

```{r}
x <- cbind(x,y) 
plot(x) 
```

The function for K-means clustering in "base" R is `kmeans()` 

```{r}
k <-kmeans(x, centers = 2) 
k 
```

To get at the results of the returned list object we can use the dollar `$` syntax 


> Q. How many points are in each cluster? 

```{r}
k$cluster 
```

>Q. What 'component' of your result object details 
  - cluster assignment/membership? 
  - cluster center? 
  
```{r}
k$centers 
```

>Q. Make a clustering results figure of the data colored by cluster membership and show cluster centers. 

```{r}
plot(x, col=c("red", "blue") ) 
```

```{r}
plot(x, col=2)
```

```{r}
plot(x, col=k$cluster, pch=16 )
points(k$centers, col="blue", pch=15, cex=2) 
```

K-means clustering is very popular as it is very fast and relatively straight forward: it takes numeric data as input and returns the cluster membership vector etc. 

The "issue" is we tell `kmeans()` how many clusters we want! 

>Q. Run kmeans again and cluster into 4 grps/clusters and plot the results like we did above? 

```{r}
k4 <- kmeans(x, centers = 4)
plot(x, col=k4$cluster) 
points(k4$centers, pch=15)
```


Scree plot to pick k `centers` value 

brute-force 
```{r}
k1 <- kmeans(x, centers=1)
k2 <- kmeans(x, centers=2)
k3 <- kmeans(x, centers=3)
k4 <- kmeans(x, centers=4)
k5 <- kmeans(x, centers=5)

```



```{r}
z <- c(k1$tot.withinss,
       k2$tot.withinss, 
       k3$tot.withinss, 
       k4$tot.withinss) 

plot(z, typ="b")
```

```{r}
n<- NULL 
for( i in 1:5) {
  n <- c(n, kmeans(x, centers=i)$tot.withinss) 
}

plot(n, typ="b")
``` 

## Hierarchical Clustering 

The main "base" R function for Hierarchical Clustering is called `hclust()`. Here we can't just input our data we need to first calculate a distance matrix (e.g. `dist()`) for our data and use this as input to `hclust()` 

```{r}
d <- dist(x) 
hc <- hclust(d)
hc 
```
There is a plot method for hclust results lets try it 

```{r}
plot(hc) 
abline(h=8, col="red")

```

To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a given height or at a height that yields a given "k" groups. 

```{r}
cutree(hc, h=8) 
```

```{r}
cutree(hc, h=8)
```

```{r}
grps <- cutree(hc, k=2) 
```

> Q. Plot the data with our hclust result coloring 

```{r}
plot(x, col=grps) 
```

# Principal Component Analysis (PCA)

## PCA of UK food data 
Import food data from an online CSV file: 

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x 
```


```{r}

x <- read.csv(url, row.names = 1)
x
```

Some base figures

```{r} 
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
There is one plot that can be useful for small datasets: 
```{r}
pairs (x, col=rainbow(nrow(x)), pch=16)
```

> Main point: It can be difficult to spot major trends and patterns even in relatively small multivariate datasets (here we only have 17 dimensions, typically we have 1000s). 

## PCA to the rescue 

The main function in "base" R for PCA is called `prcomp()` 
I will take the transpose of our data so the "foods" are in the columns: 

```{r}
pca <- prcomp( t(x) )
summary(pca)
``` 

```{r}
cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=cols, pch=16) 
```

```{r} 
#install.packages("ggplot2")
library(ggplot2)
```


```{r} 
ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point()
``` 

```{r}
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) + 
  geom_col() 
```


PCA looks super useful and we will come back to describe this further next day. 




