---
title: "Class 6: R functions"
author: "Laura (PID: A17844089)"
format: pdf
---

All functions in R have at least 3 things: 

- A **name**, we pick this and use it to call the function. 
- Input **arguments**, there can be multiple comma seperated inputs to the function. 
- The **body**, lines of R code that do the work of the function. 


Our first wee function: 

```{r}
add <- function(x, y=1) {
 x + y  
}
```

Lets test our function 

```{r}
add(10, 10)
```
```{r}
add(10)
```

## A second function 

Let's try something more interesting. Make a sequence generation tool. 

The `sample()` function could be useful here. 

```{r}
sample(1:10, size = 3) 
```
change this to work with the nucleotides A C G and T and return 3 of them 

```{r}
n <- c("A", "C", "G", "T")
sample(n, size=15, replace = TRUE)
```
Turn this snipet into a function that returns a user specified length dna sequence. Let's call it `generate_dna`... 

```{r}
generate_dna <- function(len=10, fasta=FALSE) {
  n <- c("A", "C", "G", "T")
  v <- sample(n, size=len, replace = TRUE) 
  
  # Make a single element vector 
  s <- paste(v, collapse="")
  
  cat("Well done you!\n")
  
  if(fasta) {
    return( s ) 
  } else {
    return( v )
   }
  }
``` 

```{r}
generate_dna(5)
```

```{r}
s <- generate_dna(15)
s
```

I want the option to return a single element character vector with my sequence all together like this: "GGAGTAC" 

```{r}
s
paste(s, collapse="")
``` 

I want the option to return a single element character vector with my sequence all together like this: "GGAGTAC" 
```{r}
generate_dna(10, fasta=FALSE)
```

```{r}
generate_dna(10, fasta=TRUE)
```

## A more advanced example 

Make a third function that generates protein sequence of a user specified length and format. 

```{r}
generate_protein <- function(size=15, fasta=TRUE) {
  aa <- c("A", "R", "N", "D", "C", "Q", "E", 
          "G", "H", "I", "L", "K", "M", "F",
          "P", "S", "T", "W", "Y", "V")
  
  seq <- sample(aa, size = size, replace = TRUE) 
  
  if(fasta) {
    return(paste(seq, collapse = "")) 
  } else {
    return(seq)
   }
}
``` 

Try this out... 

```{r}
generate_protein(10)
```


> Q. Generate random protein sequences between lengths 5 and 12 amino acids 

```{r}
generate_protein(5)
generate_protein(6)

```
One approach is to do this by brute force calling our function for each length 5 to 12. 

Another approach is to write a `for()` loop to iterate over the input valued 5 to 12 

A very useful third R specific approach is to use the `sapply()` function. 

```{r}
seq_lengths <- 6:12 
for (i in seq_lengths) {
  cat(">", i, "\n")
  cat( generate_protein(i) ) 
  cat("\n")
}
```

```{r}
sapply(5:12, generate_protein)
```

> **Key-Point**: Writing functions in R is doable but not the easiest thing. Starting with a working snippet of code and then using LLM tools to improve and generalize your function code is a productive approach. 



































