---
title: "Class 10: Halloween Mini Project"
author: "Laura Lu (PID: A17844089)"
format: pdf
toc: true 
---

As it is nearly Halloween and a half way point in the quarter let's do a mini project to help us figure out the best candy! 

Our come from the 538 website and is available as a CSV file: 


## Data Import 


```{r} 
candy <- read.csv( "candy-data.csv", row.names = 1)
head(candy) 
```

```{r}

flextable::flextable(head(candy,10))
```




> Q1. How many different candy types are in this dataset? 

There are 6 different candy types.


```{r} 

library(tidyverse)

candy %>%
  nrow()
```



> Q2. how many druity candy types are in the dataset? 

```{r}
sum(candy$fruity)
```

My favorite winpercent 

```{r}
 candy["Twix",]$winpercent 
```

```{r}
library(dplyr) 

candy |> 
  filter(rownames(candy)=="Almond Joy") |>
  select(winpercent)
```


## Quick overview of the dataset 


```{r}

library(skimr) 

skim(candy) 
```

```{r} 
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale tothe majority of the other columns in the dataset? 


> Q7. What do you think a zero and one represent for the candy$chocolate column? 

That the candy does not contain chocolate. 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) + 
  geom_histogram(bins=20) 
```

> Q9. Is the distribution of winpercent values symmetrical?



```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) + 
  geom_density() 
```


> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset 
# 2. Find their winpercent values 
# 3. Calculate the mean of these values 

# 4-6. Do the same for fruity candy 
# 7. Compare mean winpercents of chocolate vs fruity 
# 8. Pick the highest as the winner 

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent 
choc.mean <- mean(choc.win) 
choc.mean 
```


```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```

```{r}
mean(candy[candy$fruity==1,]$winpercent)

fruit.ind <- candy$fruity==1 
fruit.win <- candy[fruit.ind,]$winpercent 
fruit.mean <- mean(fruit.win) 
fruit.mean 

```

```{r}
candy |> 
  filter(chocolate==1) |> 
  select(winpercent) 
```


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win ) 
```


## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set? 

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x <- c(5,1,10,4) 
#sort(x)
order(x) 

#(candy$winpercent)
```

```{r}
ord.ind <- order( candy$winpercent ) 
head( candy[ord.ind,], 5) 
```

```{r}
candy |> 
  arrange(-winpercent) |> 
  head(5) 
```

> Q15. Make a first barplot of candy ranking based on winpercent values. 

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) + 
  geom_col()
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent? 

```{r}
ggplot(candy) + 
  aes(x=winpercent,
      y=reorder(rownames(candy), winpercent)) + 
  geom_col() 

```

Add some color based on the "type of candy" 

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$fruity)] <- "pink" 
my_cols[as.logical(candy$bar)] <- "purple" 
my_cols 
```

```{r}
ggplot(candy) + 
  aes(x=winpercent,
      y=reorder(rownames(candy), winpercent)) + 
  geom_col(fill=my_cols) 
```

## Winpercent and Pricepercent 

A plot with both variables/columns winpercent and pricepercent 

```{r} 
my_cols[as.logical(candy$fruity)]  <- "red"


ggplot(candy) + 
  aes(x=winpercent, 
      y=pricepercent, 
      label=rownames(candy)) + 
  geom_point(color=my_cols) + 
  geom_text() 
```

```{r} 

library(ggrepel) 


ggplot(candy) + 
  aes(x=winpercent, 
      y=pricepercent, 
      label=rownames(candy)) + 
  geom_point(color=my_cols) + 
  geom_text_repel(max.overlaps = 7) 
```

## Exploring the correlation structure 

Now that we've explored the dataset a little, we'll see how the variables interact with one another. We'll use correlation and view the results with the corrplot package to plot a correlation matrix. 

```{r} 

library(corrplot) 

cij <- cor(candy)
corrplot(cij) 

```


## Principal Component Analysis 

The function to use is called `prcomp()` with an optional `scale=T/F` argument. 

```{r}
pca <- prcomp(candy, scale=TRUE) 
summary(pca)
```

Our main PCA result figure 

```{r} 
ggplot(pca$x) + 
  aes(PC1, PC2, label=rownames(pca$x)) + 
  geom_point(col=my_cols)
  geom_text_repel(col=my_cols) 
``` 

We should also examine the variable "loadings" or contributions of the original variables to the new PCs 


```{r}
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) + 
  geom_col() 
```

```{r}
p <- ggplot(pca$x) + 
  aes(PC1, PC2, label=rownames(pca$x)) + 
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols) 
```

Interactive plots that can be zoomed on and "brushed" over can be made wit the *ploty** package. It's output is interactive and will not render to PDF 

```{r} 

library(plotly) 

#plotly(p)
```

```{r} 
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```













